---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# entsoeapi

<!-- badges: start -->
<!-- badges: end -->

The goal of entsoeapi is to create an easy wrapper for the ENTSO-E api.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("krose/entsoeapi")

```

## Security token

Read here how to get a security token.

https://transparency.entsoe.eu/content/static_content/Static%20content/web%20api/Guide.html#_authentication_and_authorisation

You should also create a `.Renviron` file in your working directory with a security token and call it GITHUB_PAT.

```
GITHUB_PAT = "<your_security_token>"

```


## Example

You use the eic codes to get the data. Let's try to find the eic code for Germany.

```{r example, warning=FALSE, message=FALSE}
library(tidyverse)
library(entsoeapi)

en_eic() %>% 
  filter(AreaTypeCode == "CTY", 
         AreaName == "Germany") %>% 
  glimpse()

```


For some of the data you need to translate the generation codes.

```{r}

en_generation_codes() %>%
  glimpse()

```

Let's get the demand in Germany.

```{r}

en_load_actual_total_load(eic = "10Y1001A1001A83F", 
                          period_start = lubridate::ymd("2020-01-01", tz = "CET"),
                          period_end = lubridate::ymd_hm("2020-01-01 23:00", tz = "CET")) %>%
  glimpse()

```

This is basically how all the functions work, so let's try to get the production data too.

```{r}

en_generation_agg_gen_per_type(eic = "10Y1001A1001A83F", 
                               period_start = lubridate::ymd("2020-01-01", tz = "CET"),
                               period_end = lubridate::ymd_hm("2020-01-01 23:00", tz = "CET")) %>%
  glimpse()


```
