---
title: "Generation Mix Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generation Mix Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include = FALSE}
run_chunks <- identical(Sys.getenv("NOT_CRAN"), "true") &&
  nchar(Sys.getenv("ENTSOE_PAT")) > 0L &&
  entsoeapi::there_is_provider()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_chunks,
  fig.width = 10,
  fig.height = 6,
  out.width = "98%",
  out.height = "98%"
)
```

```{r load-packages}
library(entsoeapi)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))
library(cli)
suppressPackageStartupMessages(library(kableExtra))
library(ggplot2)
```

## Introduction

This vignette demonstrates how to analyze electricity generation mix data using
the entsoeapi package. We'll fetch generation by production type for Germany and
create visualizations to understand the composition of the power supply.

This vignette covers:

-   Fetching generation data by production type
-   Understanding production type codes
-   Visualizing the generation mix over time
-   Calculating renewable penetration
-   Comparing different time periods

## Generation Data Overview

The entsoeapi package provides several functions for generation data:

| Function | Description | Typical Use |
|----|----|----|
| `gen_per_prod_type()` | Generation by production type | Main analysis function |
| `gen_installed_capacity_per_pt()` | Installed capacity by type | Capacity analysis |
| `gen_installed_capacity_per_pu()` | Capacity per production unit | Detailed analysis |
| `gen_per_gen_unit()` | Generation per unit | Unit-level data |
| `gen_day_ahead_forecast()` | Day-ahead forecasts | Forecasting |
| `gen_wind_solar_forecasts()` | Wind/solar forecasts | Renewable forecasting |
| `gen_storage_mean_filling_rate()` | Storage filling rates | Storage analysis |

## Fetching Generation Data

### Finding Germany's EIC Code

First, let's get Germany's bidding zone EIC code:

```{r find-german-eic}
# Germany's bidding zone EIC
de_zone <- area_eic() |>
  filter(eic_long_name == "Germany") |>
  pull(eic_code)

cli_h1("Germany Bidding Zone")
cli_text("EIC: {de_zone}")
```

### Production Type Codes

ENTSO-E uses standard production type codes:

```{r list-production-types}
asset_types |>
  filter(startsWith(x = code, prefix = "B")) |>
  select(c(code, title)) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### Fetching Generation by Production Type

Use `gen_per_prod_type()` to get generation data:

```{r fetch-generation, warning = FALSE}
cli_h1("Fetching Generation Data")

# Define time range
from_ts <- ymd(x = "2024-06-01", tz = "CET")
till_ts <- from_ts + days(7)

cli_inform("Period: {from_ts} to {till_ts}")

# Fetch generation by production type
de_generation <- gen_per_prod_type(
  eic = de_zone,
  period_start = from_ts,
  period_end = till_ts,
  tidy_output = TRUE
)

cli_alert_success("Retrieved {nrow(de_generation)} data points")
```

## Exploring the Data

### Understanding Output Columns

The output includes many columns with production type information:

```{r explore-columns}
glimpse(de_generation)
```

Key columns for generation mix analysis:

| Column                | Description                               |
|-----------------------|-------------------------------------------|
| `ts_point_dt_start`   | Timestamp                                 |
| `ts_point_quantity`   | Generation quantity                       |
| `ts_mkt_psr_type`     | Power system resource type code (B01-B20) |
| `ts_mkt_psr_type_def` | Human-readable power system resource type |

### Production Types in the Data

Check which production types are present:

```{r production-types}
cli_h1("Production Types in Dataset")

de_generation |>
  summarize(
    n_points = n(),
    total_mwh = sum(x = ts_point_quantity, na.rm = TRUE) |> round(),
    .by = c(ts_mkt_psr_type, ts_mkt_psr_type_def)
  ) |>
  arrange(desc(total_mwh)) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

## Visualization

### Stacked Area Chart of Generation Mix

Visualize the composition of the generation mix over time:

```{r stacked-area-chart}
# Prepare data for visualization
gen_plot_data <- de_generation |>
  mutate(
    ts_point_dt_start = with_tz(ts_point_dt_start, tzone = "CET"),
    date = as_date(x = ts_point_dt_start, tz = "CET"),
    hour = with_tz(time = ts_point_dt_start, tzone = "CET") |> hour()
  ) |>
  summarize(
    total_mwh = sum(ts_point_quantity, na.rm = TRUE),
    .by = c(date, hour, ts_mkt_psr_type_def)
  ) |>
  mutate(datetime = ymd_h(paste(date, hour), tz = "CET"))

# Define colors for production types
production_colors <- c(
  "Biomass" = "#FF6347",
  "Fossil Brown coal/Lignite" = "#8B4513",
  "Fossil Coal-derived gas" = "#D93333",
  "Fossil Gas" = "#4682B4",
  "Fossil Hard coal" = "#2F4F4F",
  "Fossil Oil" = "#256232",
  "Geothermal" = "#00FFFF",
  "Hydro Run-of-river head installation" = "#4169E1",
  "Hydro-electric pure pumped storage head installation" = "#1E90FF",
  "Hydro-electric storage head installation" = "#87CEEB",
  "Other renewable" = "#90EE90",
  "Other unspecified" = "#808080",
  "Solar unspecified" = "#FFD700",
  "Waste" = "#A9A9A9",
  "Wind Offshore" = "#00CED1",
  "Wind Onshore" = "#32CD32"
)

# Create stacked area plot
ggplot(
  data = gen_plot_data,
  mapping = aes(
    x = datetime,
    y = total_mwh,
    fill = ts_mkt_psr_type_def
  )
) +
  geom_area(position = "stack") +
  scale_fill_manual(
    values = production_colors,
    name = NULL
  ) +
  labs(
    title = "Germany: Daily Generation Mix Over Time (Stacked Area)",
    subtitle = "June 2024 - first 7 days",
    x = "Date",
    y = "Generation (MWh)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +
  scale_y_continuous(labels = scales::comma)
```

### Daily Average Generation by Type

Compare average daily generation by production type:

```{r daily-average}
# Calculate daily averages
daily_avg <- de_generation |>
  mutate(date = as.Date(x = ts_point_dt_start, tz = "CET")) |>
  summarize(
    avg_mw = mean(ts_point_quantity, na.rm = TRUE),
    .by = c(date, ts_mkt_psr_type_def)
  ) |>
  summarize(
    avg_daily_mwh = mean(avg_mw) * 24,
    .by = ts_mkt_psr_type_def
  ) |>
  arrange(desc(avg_daily_mwh))

# Bar chart
ggplot(
  data = daily_avg,
  mapping = aes(
    x = reorder(ts_mkt_psr_type_def, avg_daily_mwh),
    y = avg_daily_mwh
  )
) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Germany: Average Daily Generation by Production Type",
    subtitle = "June 2024",
    x = "Production Type",
    y = "Average Daily Generation (MWh)"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  coord_flip()
```

## Renewable Penetration Analysis

Calculate the share of renewable generation:

```{r renewable-penetration}
# Define renewable types
renewable_types <- c(
  "Biomass",
  "Solar unspecified",
  "Wind Offshore",
  "Wind Onshore",
  "Hydro Run-of-river head installation",
  "Hydro-electric pure pumped storage head installation",
  "Hydro-electric storage head installation",
  "Geothermal",
  "Other renewable"
)

# Calculate penetration
gen_summary <- de_generation |>
  mutate(
    is_renewable = ts_mkt_psr_type_def %in% renewable_types,
    date = as.Date(x = ts_point_dt_start, tz = "CET")
  ) |>
  summarize(
    total_gen = sum(ts_point_quantity, na.rm = TRUE),
    renewable_gen = sum(ts_point_quantity[is_renewable], na.rm = TRUE),
    renewable_pct = renewable_gen / total_gen * 100,
    .by = date
  )

print(gen_summary)

# Daily renewable share
ggplot(
  data = gen_summary,
  mapping = aes(x = date, y = renewable_pct)
) +
  geom_col(fill = "forestgreen", alpha = 0.7) +
  labs(
    title = "Germany: Daily Renewable Penetration",
    subtitle = "June 2024 - first 7 days",
    x = "Date",
    y = "Renewable Share (%)"
  ) +
  ylim(0, 100) +
  scale_x_date(date_breaks = "1 day", date_labels = "%b %d") +
  theme_minimal()
```

## Capacity Factor Analysis

Calculate capacity factors for renewable generation:

```{r capacity-factor}
# Get installed capacity (simplified - uses example values)
cli_inform(
  "Note: Full capacity analysis requires matching installed capacity data"
)

# Estimate capacity factors for solar and wind
capacity_estimates <- de_generation |>
  filter(
    ts_mkt_psr_type_def %in% c(
      "Solar unspecified", "Wind Onshore", "Wind Offshore"
    )
  ) |>
  mutate(date = as.Date(x = ts_point_dt_start, tz = "CET")) |>
  summarize(
    max_mw = max(ts_point_quantity, na.rm = TRUE),
    avg_mw = mean(ts_point_quantity, na.rm = TRUE),
    .by = c(date, ts_mkt_psr_type_def)
  ) |>
  mutate(capacity_factor = avg_mw / max_mw)

ggplot(
  data = capacity_estimates,
  mapping = aes(
    x = date,
    y = capacity_factor,
    color = ts_mkt_psr_type_def
  )
) +
  geom_line() +
  facet_wrap(~ts_mkt_psr_type_def, scales = "free_y") +
  labs(
    title = "Germany: Estimated Capacity Factors by Day",
    subtitle = "Ratio of average to peak generation",
    x = "Date",
    y = "Capacity Factor"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  ylim(0, 1)
```

## Comparing Time Periods

Compare generation mix between different time periods:

```{r compare-periods, warning = FALSE}
# Fetch data for two weeks
week1 <- gen_per_prod_type(
  eic = de_zone,
  period_start = lubridate::ymd(x = "2024-06-01", tz = "CET"),
  period_end = lubridate::ymd(x = "2024-06-08", tz = "CET"),
  tidy_output = TRUE
) |>
  mutate(period = "Week 1")

week2 <- gen_per_prod_type(
  eic = de_zone,
  period_start = lubridate::ymd(x = "2024-06-15", tz = "CET"),
  period_end = lubridate::ymd(x = "2024-06-22", tz = "CET"),
  tidy_output = TRUE
) |>
  mutate(period = "Week 3")

# Combine and compare
combined <- bind_rows(week1, week2) |>
  summarize(
    total_gen = sum(ts_point_quantity, na.rm = TRUE),
    .by = c(period, ts_mkt_psr_type_def)
  ) |>
  mutate(
    pct = total_gen / sum(total_gen) * 100,
    .by = period
  )

# Plot comparison
ggplot(
  data = combined,
  mapping = aes(x = ts_mkt_psr_type_def, y = pct, fill = period)
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Germany: Generation Mix Comparison",
    subtitle = "Week 1st vs Week 3rd of June 2024",
    x = "Production Type",
    y = "Share of Generation (%)"
  ) +
  theme_minimal()
```

## Summary

This vignette demonstrated:

1.  **Fetching generation data** with `gen_per_prod_type()`
2.  **Understanding production type codes** from B01 to B24
3.  **Visualizing the generation mix** with stacked area charts
4.  **Calculating renewable penetration** over time
5.  **Capacity factor estimation** for renewable sources
6.  **Comparing time periods** to identify trends

The entsoeapi package makes it straightforward to analyse European electricity
generation data for research, policy analysis, or business intelligence.
