---
title: "Getting Started with entsoeapi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with entsoeapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include = FALSE}
run_chunks <- identical(Sys.getenv("NOT_CRAN"), "true") &&
  nchar(Sys.getenv("ENTSOE_PAT")) > 0L &&
  entsoeapi::there_is_provider()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_chunks,
  fig.width = 10,
  fig.height = 6,
  out.width = "98%",
  out.height = "98%"
)
```

## Introduction

The `entsoeapi` package provides a standardized R interface to the ENTSO-E
(European Network of Transmission System Operators for Electricity) Transparency
Platform API. This platform publishes electricity market, generation, load,
transmission, outage, and balancing data for European countries.

This vignette will guide you through:

-   Setting up the package and obtaining API access
-   Checking connectivity and platform news
-   Finding Energy Identification Codes (EICs) for areas of interest
-   Making your first API call
-   Understanding the output structure

## Prerequisites

### Obtaining an API Token

To use the ENTSO-E Transparency Platform API, you need a security token:

1.  Register at <https://transparency.entsoe.eu/>
2.  After registration, your token should be available in your profile
3.  Set the `ENTSOE_PAT` environment variable in your .Renviron file:

```{r set-env-token, eval = FALSE}
usethis::edit_r_environ()
```

For persistent access, add this line to your `.Renviron` file:

```{bash, eval = FALSE}
# In your ~/.Renviron file:
ENTSOE_PAT=your_token_here
```

## Installation

Install the development version from GitHub:

```{r install-pkg, eval = FALSE}
# Install from GitHub
remotes::install_github("krose/entsoeapi")

# Or install from CRAN (when available)
install.packages("entsoeapi")
```

Load the packages:

```{r load-packages}
library(entsoeapi)
suppressPackageStartupMessages(library(dplyr))
library(cli)
library(lubridate)
library(kableExtra)
```

### Testing Connectivity

The `there_is_provider()` function checks if the ENTSO-E API is reachable:

```{r test-connectivity}
# Check if the API is accessible
there_is_provider()
```

### Checking Platform News

The `get_news()` function fetches the latest announcements from the ENTSO-E
Transparency Platform RSS feed. This is useful for checking planned maintenance
windows or data publication delays before running a batch of queries:

```{r check-news}
# Show the latest 3 news items
get_news(n = 3L)
```

The result is returned invisibly as a tibble, so you can also capture and filter
it:

```{r capture-news, eval = FALSE}
news <- get_news(n = 10L)
news |>
  subset(subset = grepl(pattern = "maintenance", x = title, ignore.case = TRUE))
```

## Finding EIC Codes

Energy Identification Codes (EICs) are 16-character codes that uniquely identify
market participants, bidding zones, transmission lines, and other entities on
the ENTSO-E platform.

The `all_approved_eic()` function returns a comprehensive list of approved EIC
codes:

```{r find-germany-eic}
# Find Germany's bidding zone EIC
germany_eic <- all_approved_eic() |>
  filter(eic_long_name == "Germany_Luxemburg") |>
  pull(eic_code)

cli_h1("Germany Bidding Zone EIC")
cli_text(germany_eic)
```

The `area_eic()` function provides a focused lookup for bidding zones:

```{r area-eic}
# Get all bidding zones
bidding_zones <- area_eic()
cli_h1("Available Bidding Zones")
cli_text("Total zones: {nrow(bidding_zones)}")
bidding_zones |>
  select(eic_code, eic_long_name) |>
  head(10) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

## Your First API Call

Let's fetch day-ahead energy prices for Germany. The `energy_prices()` function
queries the market data endpoint:

```{r first-api-call}
# Define time range (one week of data)
from_ts <- ymd("2026-01-01", tz = "CET")
till_ts <- from_ts + weeks(1L)

cli_h1("Querying Day-Ahead Energy Prices")
cli_inform("Period: {from_ts} to {till_ts}")
cli_inform("Area: {germany_eic}")

# Fetch the data
da_prices <- energy_prices(
  eic = germany_eic,
  period_start = from_ts,
  period_end = till_ts,
  contract_type = "A01",  # Day-ahead
  tidy_output = TRUE,
  security_token = Sys.getenv("ENTSOE_PAT")
)

cli_alert_success("Retrieved {nrow(da_prices)} rows")
```

### Understanding the Output

The `tidy_output = TRUE` format returns one row per data point:

```{r explore-output}
# Examine the structure
glimpse(da_prices)
```

Key columns in the output:

| Column                     | Description                              |
|----------------------------|------------------------------------------|
| `ts_point_dt_start`        | Timestamp for each data point            |
| `ts_point_price_amount`    | Price value                              |
| `ts_resolution`            | Time resolution (e.g., PT60M for hourly) |
| `ts_business_type`         | Type of price (Spot price = A62)         |
| `ts_market_agreement_type` | Contract type (A01 = Daily contract)     |

### The Nested Alternative

With `tidy_output = FALSE`, the data is structured differently:

```{r nested-output}
# Same query with nested output
da_prices_nested <- energy_prices(
  eic = germany_eic,
  period_start = from_ts,
  period_end = till_ts,
  contract_type = "A01",
  tidy_output = FALSE
)

glimpse(da_prices_nested)
```

With nested output, each row represents a time period, and all data points are
stored in the `ts_point` list-column. This format preserves the original API
structure and can be more efficient for certain operations.

## Common Patterns

### Setting Date Ranges

Use lubridate for robust date handling:

```{r date-ranges}
# Various date range examples
last_week <- ymd(Sys.Date()) - days(7L)
today <- ymd(Sys.Date())

# Month of data
month_start <- floor_date(Sys.Date(), "month") - months(1L)
month_end <- floor_date(Sys.Date(), "month") - days(1L)
```

### The One-Year Limit

Most ENTSO-E endpoints enforce a maximum query range of one year. The package
handles this transparently through automatic pagination:

```{r one-year-query, warning = FALSE}
# Query close to the one-year limit
year_start <- ymd("2024-01-01", tz = "CET")
year_end <- ymd("2024-12-31", tz = "CET")

# The package will handle pagination automatically if needed
# This may take longer for large datasets
cli_warn("Querying a full year - this may take a moment")

da_prices_year <- energy_prices(
  eic = germany_eic,
  period_start = year_start,
  period_end = year_end,
  contract_type = "A01",
  tidy_output = TRUE
)

cli_alert_success("Retrieved {nrow(da_prices_year)} rows for the year")
```

### Timezone Considerations

All timestamps are returned in UTC. Use lubridate to convert:

```{r timezones}
# Convert timestamps to CET/CEST
da_prices |>
  mutate(
    ts_point_dt_start_cet = with_tz(time = ts_point_dt_start, tzone = "CET")
  ) |>
  select(ts_point_dt_start, ts_point_dt_start_cet, ts_point_price_amount) |>
  head(n = 5L) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

## What's Next?

Now that you've made your first API call, explore these topics:

-   **EIC Codes**: The `area_eic()`, `tie_line_eic()`, and other EIC functions
    help you find codes for specific regions and entities
-   **Time Series Restructuring**: Understanding the `tidy_output` parameter
    helps you work with the data effectively
-   **Generation Mix**: Analyze generation by production type with
    `gen_per_prod_type()`
-   **Architecture**: See the package architecture vignette for details on how
    the package works internally
