---
title: "Working with EIC Codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with EIC Codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include = FALSE}
run_chunks <- identical(Sys.getenv("NOT_CRAN"), "true") &&
  nchar(Sys.getenv("ENTSOE_PAT")) > 0L &&
  entsoeapi::there_is_provider()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_chunks,
  fig.width = 10,
  fig.height = 6,
  out.width = "98%",
  out.height = "98%"
)
```

```{r load-packages}
library(entsoeapi)
suppressPackageStartupMessages(library(dplyr))
library(cli)
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(kableExtra))
```

## Introduction

Energy Identification Codes (EICs) are 16-character alphanumeric codes that
uniquely identify market participants, bidding zones, transmission lines, and
other entities on the ENTSO-E platform. Understanding EICs is essential for
querying data through the entsoeapi package.

This vignette covers:

-   EIC code structure and validation
-   The eight EIC lookup functions
-   Choosing between `all_approved_eic()` and `all_allocated_eic()`
-   Practical examples for finding and using EIC codes

## EIC Code Structure

### The 16-Character Format

EIC codes follow a specific format:

```         
10YDE-VE-------2
^^^^ ^^ ^^^^^^^
 ||  ||     ||
 ||  ||     ++-- Checksum (weighted-modulo-37)
 ||  ++-------- Country/area code
 ++------------- Type code
```

Example: `10YDE-VE-------2`

-   `10Y` - Type code for bidding zone (Y = area)
-   `DE` - Country code (Germany)
-   `VE` - Area code
-   `-------` - Filler characters
-   `2` - Checksum character

### EIC Type Codes

The first three characters indicate the EIC type:

| Type Code | Description                   | Example                 |
|-----------|-------------------------------|-------------------------|
| `10Y`     | Bidding zones / control areas | `10YDE-VE-------2`      |
| `10X`     | Market participants           | `10X1001A1001A42F`      |
| `10Z`     | Interconnectors / tie lines   | `10Z-DE-AT---------W`   |
| `10W`     | Power system resources        | `10WGRAR-10ZDE-ENBW--Q` |
| `10V`     | Locations                     | `10V1001A1001A48H`      |
| `10T`     | Substations                   | `10T-DE-VE-------Q`     |
| `11Y`     | Accounting points             | `11YDE-1001A0001K`      |

### Checksum Validation

The package automatically validates EIC codes using the ENTSO-E
weighted-modulo-37 algorithm. Invalid codes will produce an error:

```{r checksum-validation}
# EIC with not valid checksum - will fail
try(
  expr = energy_prices(
    eic = "ABCDEF1234567890",
    period_start = ymd("2026-01-01", tz = "CET"),
    period_end = ymd("2026-01-02", tz = "CET"),
    contract_type = "A07",
    tidy_output = TRUE
  )
)
```

## The Eight EIC Lookup Functions

The entsoeapi package provides eight functions to look up EIC codes, organized
by entity type:

### area_eic() - Bidding Zones

Returns all bidding zones and control areas:

```{r area-eic}
# Get all bidding zones
zones <- area_eic()

cli_h1("Bidding Zones")
cli_text("Total zones: {nrow(zones)}")

# Find specific countries
zones |>
  filter(
    grepl(
      pattern = "Switzerland|Schweiz|Swiss",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### party_eic() - Market Participants

Returns market participants (generators, traders, etc.):

```{r party-eic}
# Get market participants
parties <- party_eic()

cli_h1("Market Participants")
cli_text("Total parties: {nrow(parties)}")

# Find German TSOs
parties |>
  filter(
    grepl(
      pattern = "Switzerland|Schweiz|Swiss",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### accounting_point_eic() - Accounting Points

Returns accounting point EICs:

```{r accounting-point-eic}
acc_points <- accounting_point_eic()

cli_h1("Accounting Points")
cli_text("Total accounting points: {nrow(acc_points)}")

# Sample entries
acc_points |>
  filter(
    grepl(
      pattern = "Switzerland|Schweiz|Swiss",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### tie_line_eic() - Interconnectors

Returns transmission lines between bidding zones:

```{r tie-line-eic}
tie_lines <- tie_line_eic()

cli_h1("Tie Lines (Interconnectors)")
cli_text("Total interconnectors: {nrow(tie_lines)}")

# Find German interconnectors
tie_lines |>
  filter(
    grepl(
      pattern = "Switzerland|Schweiz|Swiss",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### location_eic() - Locations

Returns location EICs (V codes):

```{r location-eic}
locations <- location_eic()

cli_h1("Locations")
cli_text("Total locations: {nrow(locations)}")

locations |>
  filter(
    grepl(
      pattern = "Switzerland|Schweiz|Swiss",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### resource_object_eic() - Power Resources

Returns power system resources (generating units, loads):

```{r resource-object-eic}
resources <- resource_object_eic()

cli_h1("Power Resources")
cli_text("Total resources: {nrow(resources)}")

# Find German power plants
resources |>
  filter(
    grepl(
      pattern = "KW-DE|DE-TU",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### substation_eic() - Substations

Returns substation EICs (A codes):

```{r substation-eic}
substations <- substation_eic()

cli_h1("Substations")
cli_text("Total substations: {nrow(substations)}")

# Sample entries
substations |>
  head(12L) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### all_approved_eic() - All Approved EICs

Combines all approved EICs into a single tibble:

```{r all-approved-eic}
all_eic <- all_approved_eic()

cli_h1("All Approved EICs")
cli_text("Total EICs: {nrow(all_eic)}")

# Count by type
all_eic |>
  count(type, sort = TRUE) |>
  mutate(pct = round(x = n / sum(n) * 100, digits = 2L)) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

## all_approved_eic() vs all_allocated_eic()

The package provides two comprehensive EIC functions with different
characteristics:

### Comparison Table

| Feature          | all_approved_eic() | all_allocated_eic() |
|------------------|--------------------|---------------------|
| Data source      | CSV downloads      | XML download        |
| Speed            | Fast (seconds)     | Slow (minutes)      |
| Update frequency | Less frequent      | More current        |
| Columns          | Standardized       | Extended details    |
| Columns returned | 11 columns         | 18 columns          |

### all_approved_eic()

Fast and efficient for most use cases:

```{r all-approved-detail}
cli_h1("all_approved_eic() - Column Details")
approved_eic <- all_approved_eic()
glimpse(approved_eic)
```

Use this function when you need:

-   Quick lookups
-   Basic EIC information
-   Standard column structure

### all_allocated_eic()

Provides more detailed information but is slower:

```{r all-allocated-detail, warning = FALSE}
cli_h1("all_allocated_eic() - Column Details")
cli_inform("Note: This function is slow (downloads ~70MB XML)")

allocated_eic <- all_allocated_eic()
glimpse(allocated_eic)
```

Column comparison between `all_allocated_eic()` and `all_approced_eic`:

```{r}
cli_h1("all_allocated_eic()")
allocated_eic |>
  filter(eic_code == "50WG00000001997X") |>
  t() |>
  kbl(format = "pipe") |>
  cat(sep = "\n")

cli_h1("all_approved_eic()")

approved_eic |>
  filter(eic_code == "50WG00000001997X") |>
  t() |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

Use `all_allocated_eic()` function when you need:

-   More detailed participant information
-   Current allocation status
-   Extended metadata

## Practical Examples

### Finding Germany's EIC Codes

```{r find-germany-codes}
cli_h1("German bidding zones")
all_eic |>
  filter(
    grepl(
      pattern = "Germany.*bidding|area.*germany",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")

cli_h1("German TSOs")
party_eic() |>
  filter(
    grepl(
      pattern = "50Hertz|Amprion|TenneT|TransnetBW",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  filter(
    grepl(
      pattern = "System Operator",
      x = eic_type_function_list,
      ignore.case = TRUE
    )
  ) |>
  filter(
    grepl(
      pattern = "GmbH",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### Finding Nordic Bidding Zones

```{r nordic-zones}
cli_h1("Nordic Bidding Zones")

nordic_countries <- c("Sweden", "Norway", "Denmark", "Finland")
zones |>
  filter(eic_long_name %in% nordic_countries) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### Finding Transmission Lines Between Countries

```{r cross-border}
cli_h1("Cross-Border Transmission Lines")

# Find Germany-France interconnector
tie_line_eic() |>
  filter(
    grepl(
      pattern = "DE.*FR|FR.*DE|Germany.*France",
      x = eic_long_name,
      ignore.case = TRUE
    )
  ) |>
  select(eic_code, eic_long_name, type) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

### Combining EIC Data with API Queries

Use EIC lookups to construct queries:

```{r combine-with-query, warning = FALSE}
cli_h1("Using EICs in Queries")
# Query intraday prices for Spain
from_ts <- ymd("2026-01-01", tz = "CET")
till_ts <- from_ts + days(3L)

es_prices <- energy_prices(
  eic = "10YES-REE------0",
  period_start = from_ts,
  period_end = till_ts,
  contract_type = "A07",
  tidy_output = TRUE
)

cli_alert_success("Retrieved {nrow(es_prices)} intraday price points for Spain")

# Show sample data
es_prices |>
  filter(ts_classification_sequence_position == 1L) |>
  select(
    market_agreement_type_def,
    ts_point_dt_start,
    ts_point_price_amount,
    ts_currency_unit_name
  ) |>
  head(12) |>
  kbl(format = "pipe") |>
  cat(sep = "\n")
```

## Summary

The entsoeapi package provides eight EIC lookup functions to help you find the
codes needed for your queries:

| Function                 | Returns             | EIC Type  |
|--------------------------|---------------------|-----------|
| `area_eic()`             | Bidding zones       | Y codes   |
| `party_eic()`            | Market participants | X codes   |
| `accounting_point_eic()` | Accounting points   | Y codes   |
| `tie_line_eic()`         | Interconnectors     | Z codes   |
| `location_eic()`         | Locations           | V codes   |
| `resource_object_eic()`  | Power resources     | W codes   |
| `substation_eic()`       | Substations         | A codes   |
| `all_approved_eic()`     | All approved EICs   | All types |

For comprehensive lookups, use `all_approved_eic()` for speed or
`all_allocated_eic()` for extended details.
